Figures to insert in text

### 2018-03-08 Chapters/Chapter02.tex
SDM Performance data for Amazon EC2 p2.xlarge and p3.2xlarge.

Murilo observed that the models of Kanerva's read ($z=1$) and Chada's read ($z=0$) were simple variations of a generalized read with an exponent $z$, which suggests experimenting with different values. Mathematically, let $A$ be the set of the counters of the activated hardlocation, and $c_i$ be the counter of the $i$-th bit. Then,

$$
s_i = \sum_{c \in A} \frac{c_i}{|c_i|} |c_i|^z
$$

The sum of $|c_i|^z$ turns the intermediate values from integers to float point numbers. Thus, we have developed a specific read operation which stored the intermediate values in double variables.

The results, however, have not yielded performance improvements. Though for $z \leq 1$ results are comparable to $z=1$, for $z>1$, the system shows a clear deterioration, with a smaller critical distance and faster divergence at large-distance reads. This is shown in Figures \ref{fig:murillo-generalization-experiments} and \ref{fig:murillo-generalization-experiments-6reads}.

We understand that the critical distant is an important parameter of SDM. The bigger the critical distance, the best, because SDM is able to converge even with farther clues. For $z>1$, the bigger the $z$, the smaller the critical distance. For $z = 6$, the critical distance almost reaches zero.

It is interesting that Kanerva has proposed $z=1$ without realizing the generalized reading. Even so, he proposed the optimal $z$ --- the one with the highest critical distance.

@@

  \caption{(a) and (b) show the behavior of a single read. As stated previously, we can see a deterioration of convergence, with lower critical distance as $z>1$.  Another observation can be made here, concerning the discrepancy of Kanerva's Fig 7.3 and our data.  It seems that Kanerva may not have considered that a single read would only `clean' a small number of dimensions \emph{after the critical distance}. What we observe clearly is that with a single read, as the distance grows, the system only `cleans' towards the orthoghonal distance 500 after a number of iterative readings.}
  \label{fig:murillo-generalization-experiments}
\end{figure}

\begin{figure}[h!]
  \centering
  \subfloat[$z \in \{0, 1\}$]{\includegraphics[width=\textwidth]{./images02/new-images/z_0_1.png}}

  \subfloat[$z \in \{0, 0.5, 1, 1.5, 3, 4.5, 6\}$]{\includegraphics[width=\textwidth]{./images02/new-images/iter_z_all_2.png}}

  \caption{(a) and (b) show the behavior of Figure \ref{fig:murillo-generalization-experiments}, now executed with 6-iterative reads. What we observe clearly is that with a single read, as the distance grows, the system only `cleans' towards the orthoghonal distance 500 after a number of iterative readings.}
  \label{fig:murillo-generalization-experiments-6reads}
\end{figure}

@@

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 100 & 8.59 & 116.38 \\
    Thread scan & 500 & 18.66 & 53.56 \\
    OpenCL scan & 1000 & 77.20 & 12.95 \\
    \hline
\end{tabular}
\caption{iMac Retina 5K 27-inch 2017 with a 3.8GHz Intel core i5 processor, 8GB DDR4 RAM, and a Radeon Pro 580 8G GPU. Running an SDM with $n=10,000$ bits, $H=1,000,000$, and $r=4845$.  There is no benchmark for read and write operations because RAM is not enough to allocate the counters --- it would consume 37.25 GB of RAM.
\label{tab:perf-imac-10k}}
\end{table}

@@

Deep neural networks, alphaGo, alphazero

@@

A reduction from measuring the distance to $2^{20}$ hard locations to a computation of $2^{10}$ hard locations might yield astonishing performance gains, depending, of course, on our optimistic assumptions concerning existence and complexity of such algorithm.  At large scales of computing, the very ability to perform some experiments is a function of sheer performance. The horizon of experiments --- and possibly of knowledge --- expands \emph{as a function of computational demands}. A little more on this in my closing words.

@@

The overarching intention here is to not only provide a starting point, but to provide a Framework in which SDM research can be conducted.  Consider, for example, having the ability to compare the results of a new (‘forked’) model to the previous `best' (under a particular benchmark set).  For example, some of the benchmarks that we plan to develop in future research is: how fast is convergence through iterative reading?  How large is the attractor of the critical distance?  How well does the system filter noise?  How well does the system work under the supervised learning task?  And other authors may be able to improve this benchmark set themselves, as is usual in open source development.  It is perhaps this facility of ease to build on top of previous work that seems most exciting at this stage.

Consider the misunderstanding concerning the SDM read operation:  Dr Stan Franklin describes Kanerva's read operation in a way that each hard location, at each dimension, provides only a single bit of information to the read operation (instead of Kanerva's full counter).  We have referred to this modified read operation as Chada read (the tale is that my friend \text{\&} colleague, Dr. Daniel de Magalhães Chada, along with Linhares, did not consult Kanerva's book and only discovered the discrepancy in code and ideas a couple of years afterwards).  Having an open, testable, codebase reduces the possibilities of such misunderstandings in the long run.  Indeed, a high-quality codebase seems to have become a scientific community's form of unequivocally standing behind a consensus. For example, the journal Nature analysed the top-100 cited papers in history, to find:

\begin{quote}
... some surprises, not least that it takes a staggering 12,119 citations to rank in the top 100 — and that many of the world’s most famous papers do not make the cut. A few that do, such as the first observation of carbon nanotubes (number 36) are indeed classic discoveries. But the vast majority describe experimental methods or \emph{software that have become essential in their fields. [...] The list reveals just how powerfully research has been affected by computation and the analysis of large data sets.} \\
\hfill --- \citet{van2014top}, emphasis mine.
\end{quote}

It is no coincidence that scientific journals such as BMC Neuroscience, or the Journal of Machine Learning Research have specific sections on open-source software. Neurocomputing states that “software is scientific method by machine”.

Of course, for the skeptical reader who may consider software a less worthy pursuit, there is also new work here.  The mathematics of the model has been shown to be correct numerically (with a single, small, anomaly); we have shown how to execute unsupervised learning with nothing besides operations original to the SDM; we have studied the generalized Murilo read; we have seen noise filtering; the death of neurons; how information-theory may be of use; and finally, we have reproduced numerous of the original propositions put forth by Kanerva.  The emphasis might have been on \emph{breadth of topics}, in detriment of depth here or there.  But this is due to our research group's enthusiam for the topic; we do indeed believe that SDM is --- if not correct --- extremely close to a full scientific understanding of human long-term memory.  If so, it is such a monumental achievement that we want readers to be able to see all of what we see and imagine the vastness of possibilities.  The work on, say, reinforcement learning, is most definitely not the definitive work we will see on the subject, but a challenge left for readers to contemplate. Some essays completely shut the door close at the end; this one intends to leave it wide open. As the reader may have noticed, this final section does not read as an analysis of work previously done; it reads, instead, as a \emph{desiderata}, a prologue, a yearning for others to join me in imagining the shape of things to come.

\begin{figure}[!p]
\centering

\subfloat[]{\includegraphics[width=0.5\textwidth]{images02/new-images/Only-one-life077.jpeg}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{images02/new-images/Only-one-life078.jpeg}}

\subfloat[]{\includegraphics[width=0.5\textwidth]{images02/new-images/Only-one-life079.jpeg}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{images02/new-images/Only-one-life080.jpeg}}

\subfloat[]{\includegraphics[width=0.5\textwidth]{images02/new-images/Only-one-life081.jpeg}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{images02/new-images/Only-one-life082.jpeg}}

\caption{Some slides from an old Linhares class \citet{only-one-life}.  In the history of computing business, the most interesting level of analysis seems to be that of the \emph{platform}.  It is where things can get built on top of, and communities emerge, and standards fight against each other, and fortunes are built or lost.  Each time a new computing platform appears, it seems as the opportunities are ripe for the taking; as if a multitude of doors have opened simultaneously.
\label{fig:one-life}}
\end{figure}

>>>>

In this chapter we will explore how Murilo's generalized read operation works and whether it may improve SDM behavior. In this case, we understand that SDM behavior will have been improved when the critical distance increases, which means it is able to recognize things with even less clues than before.

@@

  \caption{(a) and (b) show the behavior of a single read; As stated previously, we can see a deterioration of convergence, with lower critical distance as $z>1$.  Another observation can be made here, concerning the discrepancy of Kanerva's Fig 7.3 and our data.  It seems that Kanerva may not have considered that a single read would only `clean' a small number of dimensions \emph{after the critical distance}. What we observe clearly is that with a single read, as the distance grows, the system only `cleans' towards the orthoghonal distance 500 after a number of iterative readings.}
  \label{fig:murillo-generalization-experiments}
\end{figure}

\begin{figure}[h!]
  \centering
  \subfloat[$z \in \{0, 1\}$]{\includegraphics[width=3.2in]{./images02/new-images/z_0_1.png}}

  \subfloat[$z \in \{0, 0.5, 1, 1.5, 3, 4.5, 6\}$]{\includegraphics[width=3.2in]{./images02/new-images/iter_z_all_2.png}}

  \caption{(a) and (b) show the behavior of the previous Figure, now executed with 6 iterative reads. What we observe clearly is that with a single read, as the distance grows, the system only `cleans' towards the orthoghonal distance 500 after a number of iterative readings.}
  \label{fig:murillo-generalization-experiments-6reads}
\end{figure}

Murilo observed that the models of Kanerva-read ($z=1$) and Chada-read ($z=0$) were simple variations of the exponent $z$, which suggests experimenting with different values. The results, however, have not yielded performance improvements.  Though for $z \leq 1$ results are comparable to $z=1$, for $z>1$, the system shows a clear deterioration, with a smaller distance to convergence and higher divergence at large-distance reads. This is shown in Figures \ref{fig:murillo-generalization-experiments} and \ref{fig:murillo-generalization-experiments-6reads}.

@@

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 100 & 8.59 & 116.38 \\
    Thread scan & 500 & 18.66 & 53.56 \\
    OpenCL scan & 1000 & 77.20 & 12.95 \\
    \hline
\end{tabular}
\caption{iMac Retina 5K 27-inch 2017 with a 3.8GHz Intel core i5 processor, 8GB DDR4 RAM, and a Radeon Pro 580 8G GPU. Running an SDM with $n=10,000$ bits, $H=1,000,000$, and $r=4845$.  There is no benchmark for read and write operations because RAM was not enough to allocate the counters --- it would consume 37.25 GB of RAM.
\label{tab:perf-imac-10k}}
\end{table}

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 1000 & 15.44 & 64.73 \\
    Thread scan & 5000 & 30.47 & 32.81 \\
    OpenCL scan & 5000 & 93.63 & 10.67 \\ \hline
    \hline
    & Loops & Ops / second & Op. time (ms) \\ \hline
    Thread write & 1000 & 23.44 & 42.64 \\
    Thread single read & 1000 & 28.26 & 35.37 \\
    OpenCL write & 1000 & 51.78 & 19.31 \\
    OpenCL single read & 1000 & 85.31 & 11.72 \\
    \hline
\end{tabular}
\caption{Amazon EC2 p2.xlarge with Intel Xeon E5-2686v4 processor, 61GB DDR3 RAM, and NVIDIA K80 GPU. Running an SDM with $n=1,000$ bits, $H=1,000,000$, and $r=451$.}
\end{table}

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 1000 & 52.35 & 19.09 \\
    Thread scan & 5000 & 100.48 & 9.95 \\
    OpenCL scan & 5000 & 145.18 & 6.88 \\ \hline
    \hline
    & Loops & Ops / second & Op. time (ms) \\ \hline
    Thread write & 2000 & 84.73 & 11.80 \\
    Thread single read & 2000 & 95.26 & 10.49 \\
    OpenCL write & 2000 & 113.01 & 8.84 \\
    OpenCL single read & 2000 & 133.19 & 7.50 \\
    \hline
\end{tabular}
\caption{Amazon EC2 p2.xlarge with Intel Xeon E5-2686v4 processor, 61GB DDR3 RAM, and NVIDIA K80 GPU. Running an SDM with $n=256$ bits, $H=1,000,000$, and $r=103$.}
\end{table}

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 100 & 1.66 & 600.75 \\
    Thread scan & 500 & 3.37 & 296.56 \\
    OpenCL scan & 1000 & 21.39 & 46.73 \\ \hline
    \hline
    & Loops & Ops / second & Op. time (ms) \\ \hline
    Thread write & 500 & 2.60 & 383.50 \\
    Thread single read & 500 & 3.24 & 307.97 \\
    OpenCL write & 1000 & 8.18 & 122.17 \\
    OpenCL single read & 55.47 & 18.02 & 55.47 \\
    \hline
\end{tabular}
\caption{EC p2.xlarge 10,000-bits}
\caption{Amazon EC2 p2.xlarge with Intel Xeon E5-2686v4 processor, 61GB DDR3 RAM, and NVIDIA K80 GPU. Running an SDM with $n=10,000$ bits, $H=1,000,000$, and $r=4845$.}
\end{table}

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 1000 & 17.13 & 58.34 \\
    Thread scan & 5000 & 60.98 & 16.39 \\
    OpenCL scan & 5000 & 1175.88 & 0.85 \\ \hline
    \hline
    & Loops & Ops / second & Op. time (ms) \\ \hline
    Thread write & 1000 & 35.12 & 28.47 \\
    Thread single read & 1000 & 57.31 & 17.44 \\
    OpenCL write & 1000 & 113.98 & 8.77 \\
    OpenCL single read & 1000 & 549.12 & 1.82 \\
    \hline
\end{tabular}
\caption{Amazon EC2 p3.2xlarge with Intel Xeon E5-2686v4 processor, 488GB DDR3 RAM, and NVIDIA Tesla V100 GPU. Running an SDM with $n=1,000$ bits, $H=1,000,000$, and $r=451$.}
\end{table}

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 1000 & 56.80 & 17.60 \\
    Thread scan & 5000 & 192.54 & 5.19 \\
    OpenCL scan & 5000 & 1565.28 & 0.63 \\ \hline
    \hline
    & Loops & Ops / second & Op. time (ms) \\ \hline
    Thread write & 2000 & 131.64 & 7.59 \\
    Thread single read & 2000 & 161.97 & 6.17 \\
    OpenCL write & 2000 & 427.38 & 2.33 \\
    OpenCL single read & 2000 & 987.95 & 1.01 \\
    \hline
\end{tabular}
\caption{Amazon EC2 p3.2xlarge with Intel Xeon E5-2686v4 processor, 488GB DDR3 RAM, and NVIDIA Tesla V100 GPU. Running an SDM with $n=256$ bits, $H=1,000,000$, and $r=103$.}
\end{table}

\begin{table}[!htb]
\centering
\begin{tabular}{| l | r | r | r |}
    \hline
    & Loops & Scans / second & Scan time (ms) \\ \hline
    Linear scan & 100 & 1.84 & 540.97 \\
    Thread scan & 500 & 5.03 & 198.74 \\
    OpenCL scan & 1000 & 174.05 & 5.74 \\ \hline
    \hline
    & Loops & Ops / second & Op. time (ms) \\ \hline
    Thread write & 500 & 4.50 & 222.08 \\
    Thread single read & 500 & 6.89 & 145.01 \\
    OpenCL write & 1000 & 12.42 & 80.48 \\
    OpenCL single read & 1000 & 71.46 & 13.99 \\
    \hline
\end{tabular}
\caption{Amazon EC2 p3.2xlarge with Intel Xeon E5-2686v4 processor, 488GB DDR3 RAM, and NVIDIA Tesla V100 GPU. Running an SDM with $n=10,000$ bits, $H=1,000,000$, and $r=4845$.}
\end{table}

@@

@@

A reduction from measuring the distance to $2^{20}$ hard locations to a computation of $2^{10}$ hard locations might yield astonishing performance gains, depending, of course, on our optimistic assumptions concerning existence and complexity of such algorithm.  At large scales of computing, the very ability to perform some experiments is a function of sheer performance. The horizon of experiments --- and possibly of knowledge --- expands \emph{as a function of computational demands}. More on this in my closing words.

@@

It is no coincidence that a scientific journal such as Neurocomputing states that “software is scientific method by machine”.  Indeed, recently, Nature analysed the top-100 cited papers in history, to find

\begin{quote}
... some surprises, not least that it takes a staggering 12,119 citations to rank in the top 100 — and that many of the world’s most famous papers do not make the cut. A few that do, such as the first observation of carbon nanotubes (number 36) are indeed classic discoveries. But the vast majority describe experimental methods or software that have become essential in their fields. [...] The list reveals just how powerfully research has been affected by computation and the analysis of large data sets. \\
\hfill --- \citet{van2014top}
\end{quote}

The overarching intention here is to not only provide a starting point, but to provide the Framework in which SDM research can be conducted.  Consider, for example, having the ability to compare the results of a new (‘forked’) model to the previous best, under a particular benchmark set.  For example, some of the benchmarks that we plan to develop for future research is: how fast is convergence through iterative reading?  How large is the attractor of the critical distance?  How well does the system filter noise?  How well does the system work under the aforementioned supervised learning task?  And other authors may be able to improve this benchmark set themselves, as is usual in open source development.  It is perhaps this facility of ease to build on top of previous work that seems most exciting at this stage.

Consider the misunderstanding concerning the SDM read operation:  Dr Stan Franklin describes Kanerva's read operation in a way that each hard location, at each dimension, provides only a single bit of information to the read operation (instead of Kanerva's full counter).  We have referred to this modified read operation as Chada read.  Having an open, testable, codebase reduces the possibilities of such misunderstandings in the long run.

Of course, there is also new work here.  The mathematics of the model has been shown to be correct (with a single exception); we have shown how to execute unsupervised learning with nothing besides operations originally proposed by Kanerva; we have studied the generalized Murilo read; we have seen and finally, we have reproduced

\subsection{From theory to a platform.}

Platforms in the history of computing.

Opens 10 doors.
